---
title: "Demography and Event History Analysis using R"
author: "Göran Broström"
date: '2015-06-18'
output:
  slidy_presentation:
    highlight: tango
  html_document: default
  ioslides_presentation:
    fig_height: 3
    highlight: tango
    logo: figures/umu15.png
  pdf_document: default
  beamer_presentation:
    fig_height: 3
    highlight: tango
    keep_tex: yes
css: style.css
---

##Introduction


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, cache = TRUE, warnings = FALSE)
table <- function(..., useNA = 'ifany') base::table(..., useNA = useNA)
```

- Read *INDIVIDUAL.txt* and *INDIV_INDIV.txt* into **R**.
- Perform some selections and transformations.
- Save into **R** format, *.rda* files

## Reading the INDIVIDUAL file


```{r}
options(stringsAsFactors = FALSE)
individual <- read.table("rawdata/INDIVIDUAL.txt", 
                         header = FALSE)
names(individual) <- c("Id", "Id_D", "Id_I", "Source", 
                 "Type", "Value", "Value_Id_C",
                 "Day", "Month", "Year",
                 "Start_day", "Start_month", "Start_year",
                 "End_day", "End_month", "End_year",
                 "DateOfOccurrence", "TimeInvariant")
```

## Looking at the variable names ("Type")

<tiny>

```{r echo = FALSE}
x <- matrix(sort(unique(individual$Type)), ncol = 4)
colnames(x) <- c("1", "2", "3", "4")
knitr::kable(x)
```

</tiny>

- Most of these variables can be thrown out.
- Keep: BIRTH_DATE, BIRTH_LOCATION, DEATH_DATE, END_OBSERVATION, LEGITIMACY, MULTIPLE_BIRTH, OCCUPATION, START_OBSERVATION, SEX

## The R package 'dplyr' (Hadley Wickham)


```{r}
library(dplyr)
individual <- filter(individual, Type %in% c("BIRTH_DATE", "DEATH_DATE", "START_OBSERVATION", "END_OBSERVATION", "LEGITIMACY", "MULTIPLE_BIRTH", "OCCUPATION_HISCO", "SEX", "MARRIAGE_DATE", "BIRTH_LOCATION"))
```

- **filter** selects *observations* (rows) by a *condition*
- filter(data, condition)
-  **x %in% y** returns *a logical vector* of the same length as *x*:
```{r}
c(1, 3, 2, 5) %in% c(1, 2)
```

