---
title: "Using R"
author: "Göran Broström"
date: '2016-11-30'
output:
  html_document: default
  beamer_presentation:
    fig_height: 3
    highlight: tango
    keep_tex: yes
  ioslides_presentation:
    fig_height: 3
    highlight: tango
  pdf_document: default
  slidy_presentation:
    highlight: tango
css: style.css
---

```{setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(cache=TRUE,echo=TRUE,message=FALSE,warning=FALSE,fig.height = 4, fig.path="figslides/")
table <- function(..., useNA = 'ifany') base::table(..., useNA = useNA)
library(knitr)
library(dplyr)
```

## What is R?

**R** is

-  a language and environment for *statistical computing* and *graphics*.
-  a *GNU project* originating from the *S* language and
environment which was developed at *Bell Laboratories* (formerly AT&T, now
Lucent Technologies) by *John Chambers* and colleagues.

-  a different implementation of S.
-  25 years old.




From [The R project](http://www.r-project.org/about.html)


## What is R?


-  R provides a wide variety of statistical and graphical techniques,
  and is *highly extensible*.
-  One of R’s strengths is the ease with which *well-designed
  publication-quality plots* can be produced, including mathematical symbols
  and formulae where needed.


## The R environment

**R** includes

-     an effective *data handling* and *storage* facility,
-     operators for *calculations on arrays*, in
  particular *matrices*,
-     intermediate
  *tools for data analysis*,
-     superb *graphical facilities*, and
-     a simple and effective *programming language*
  which includes

    -  *conditionals*,
    -  *loops*,
    -  user-defined recursive functions and
    -  *input* and *output* facilities.

## Why R?

-  **R** is *free*:
    -  as in *free beer*,
    -  as in *free speech*.

-  **R** can be installed on as many computers you like
-  **R** is *open-source*.
-  **R** has thousands (7000+ today) of *add-on packages*.


## What is RStudio?

RStudio allows you to *run R* in a more *user-friendly* environment


-  It is *free* and *open-source*,
-  Tools for *interacting* statistical *analysis* with report
  *writing*.
    -  *R markdown*
    -  *knitr*
    -  LaTeX
-  Tools for *version control* (git, gitHub)
-  [Support](http://www.rstudio.com).


## Why RStudio?

Also,

-  It is *available* on all (Linux, Mac, Windows) platforms.
-  It looks *exactly the same* on all platforms.
      -  It *behaves exactly the same* on all platforms.

## What is reproducible research?

-  It can be *reproduced by others*,
-  It can be reproduced *by yourself* (6 months later)!


Focus is on reproducing results *from a single set of observations*.


## What is reproducible research *not*?

-  We do  *not* mean *replicable*.
    - **Replicable:** If the *results* can be replicated on
      *another* study population.
-  It is rather about rules of *disciplin* and *documentation*.
-  *RStudio* gives you the *tools*.


## Version control

-  Must be *easy* to go back to *earlier versions* of your research:
    -  You detect *errors*,
    -  You get *new data*,
    -  The journal asks you to perform *separate analyses* for men
      and women,
      -  etc.
-  There are *tools* to help:
    -  [git](http://git-scm.com), by Linus Torvalds,
        -  [gitHub](https://github.com)


## Analysis

I want to study infant mortality with data from the file *observation*. So I load it, and also the file *person*.

```{r}
load("data/observation.rda")
load("data/person.rda")
```

There are some redundant information in *observation* which we remove:

```{r}
names(observation)
```

## Load dplyr

```{r}
library(dplyr)
obs <- select(observation, idi, birthdate = birth_date, sex, occupation, married, enter, exit, event, m_idi)
summary(obs)
```
## Glitches

There are some glitches here: A quick fix:

```{r}
obs$enter <- pmax(obs$enter, 0)
with(obs, sum(exit - enter <= 0))
```

Good!

## The occupation:

```{r}
obs <- filter(obs, occupation < 100000)
obs$occ <- cut(obs$occupation, seq(1:10) * 10000, right = FALSE, 
               labels = 1:9)
with(obs,  table(event, occ))
```

## First survival analysis

<small>
```{r}
library(eha)
fit <- coxreg(Surv(enter, exit, event) ~ occ, data = obs)
##drop1(fit, test = "Chisq")
```
</small>

## The coefficients

```{r}
summary(fit)
```

## Mother's age

```{r}
obs$birthdate <- as.numeric(toTime(obs$birthdate))
indx <- match(obs$m_idi, person$idi)
obs$m.birthdate <- person$birth_date[indx] 
obs$m.birthdate <- as.numeric(toTime(obs$m.birthdate))
obs$m.age <- with(obs, birthdate + enter - m.birthdate)
select(obs, birthdate, idi, m_idi, m.age, enter, exit) %>% head(3)
```

## Infant mortality

```{r}
source("R/ageWin.R")
infant <- age.window(as.data.frame(obs), c(0, 1))
summary(infant)
```

## Cox regression

```{r}
fit <- coxph(Surv(enter, exit, event) ~ m.age, data = infant)
summary(fit)
```

## Cumulative hazards

```{r}
plot(survfit(fit), fun = "cumhaz", mark.time = FALSE)
```


