---
title: "Convert IDS files to R format"
author: "Göran Broström"
date: "November 23, 2016"
output:
  html_document: default
  pdf_document: default
  word_document: default
bibliography: ids.bib
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, cache = TRUE, warnings = FALSE)
table <- function(..., useNA = 'ifany') base::table(..., useNA = useNA)
```

## Introduction

The two IDS data sets we will use are stored in ordinary text files. It is convenient to convert these files to the *internal format of* **R**, files with the extension *.rda*. In the process of doing this, we also do some selections and transformations. Most importantly, we split the file *INDIVIDUAL* into two parts, *person* and *observation*. *person* contains the *time-fixed* information and *observation* contains the *time-varying*.

The raw data files, *INDIVIDUAL.txt* and *INDIV_INDIV.txt* are stored in the subdirectory *rawdata*, and the converted files, *person.rda*, *observation.rda* and *INDIV_INDIV.rda* will be stored in the subdirectory *data*. 

## The INDIVIDUAL file

In reading textfiles into **R**, the standard tool is the function *read.table*:

```{r readrawdata}
options(stringsAsFactors = FALSE)
individual <- read.table("rawdata/INDIVIDUAL.txt", header = FALSE)
names(individual)[1:7] <- c("Id", "Id_D", "Id_I", "Source", 
                            "Type", "Value", "Value_Id_C")
names(individual)[8:10] <- c("Day", "Month", "Year")
names(individual)[11:13] <- c("Start_day", "Start_month", "Start_year")
names(individual)[14:16] <- c("End_day", "End_month", "End_year")
names(individual)[17:18] <- c("DateOfOccurrence", "TimeInvariant")
```

This data file is heavily *un-tidy* [@tidy-data]. *Tidy data* have variables in columns and observations in rows. The column *Type* in *INDIVIDUAL* has the following collection of possible values (with frequencies):

```{r showType}
with(individual, table(Type))
```

For our small example, we select just a *few Type*s. The package *dplyr* is very handy for data manipulation in *pseudo-SQL* style. We start by loading it, and then we use the function *filter*, which select *observations* (rows) by certain conditions.

```{r selectTypes}
library(dplyr)
individual <- filter(individual, Type %in% c("BIRTH_DATE", "DEATH_DATE",
                                             "START_OBSERVATION", "END_OBSERVATION",
                                             "LEGITIMACY", "MULTIPLE_BIRTH",
                                             "OCCUPATION_HISCO", "SEX", 
                                             "MARRIAGE_DATE", "BIRTH_LOCATION"))
```

We first look at the column *TimeInvariant*:

```{r timeInvariant}
select(individual, Type, TimeInvariant) %>% table()
```

The symbol "*%>%*"" is a *pipe*: You can think of it as saying "*and then*". It is equivalent to a two-step nested procedure:

```{equi, eval = FALSE}
table(select(individual, Type, TimeInvariant))
```

or non-nested:

```{r non-nest, eval = FALSE}
x <- select(individual, Type, TimeInvariant)
table(x)
rm(x) # Remove the temporary variable 'x'
```

The advantages with the pipe is:

- It is easy to read (once you are familiar with the pipe concept).

- It does not clutter with temporary variables: All 'dirty work' is hidden.

- It is *fast*: The *dplyr* package is written in optimized *C++* code.

Obviously, the variable *TimeInvariant* takes the values "" (the empty string) and "TimeInvariant". Not very logical. Furthermore, the (only) variables that are *truly time-invariant* are *BIRTH_DATE*, *BIRTH_LOCATION*, *DEATH_DATE*, *LEGITIMACY*, *MULTIPLE_BIRTH*, and *SEX*.
So we recode, relabel, and convert to *logical*:

```{r recode TimeInvariant}
individual$TimeInvariant <- with(individual, 
                                 Type %in% c("BIRTH_DATE", "BIRTH_LOCATION",
                                             "DEATH_DATE", "LEGITIMACY",
                                             "MULTIPLE_BIRTH", "SEX"))
select(individual, Type, TimeInvariant) %>% table()
```

This is more like it. Now the simple continuation is to split the data frame *individual* into two according to whether "TimeInvariant" is TRUE or FALSE. Let us call the two data frames 
*person* and *observation*, respectively:

```{r splitfile}
person <- filter(individual, TimeInvariant)
person$TimeInvariant <- NULL # Removes 'TimeInvariant' from 'person'
observation <- filter(individual, !TimeInvariant) # '!' reads 'NOT', logical negation
observation$TimeInvariant <- NULL

```

Now let us look at the column *DateOfOccurrence* in the two data frames (I use the "standard R" way of using *table* here, as a comparison):

```{r selectedTypes}
with(person, table(Type, DateOfOccurrence))
with(observation, table(Type, DateOfOccurrence))
```

Note that *DateOfOcurrence* takes two values, "" (empty string) and "Date of occurrence". More logical is to let it take the values *FALSE* and *TRUE*, respectively:

```{r recodeDOO}
person$DateOfOccurrence <- person$DateOfOccurrence != ""
with(person, table(Type, DateOfOccurrence))
observation$DateOfOccurrence <- observation$DateOfOccurrence != ""
with(observation, table(Type, DateOfOccurrence))
```

We draw the conclusion that the variable *DateOfOccurrence* is redundant in both *person* and *observation*, so we remove it. And we remove the data frame *individual* at the same time:

```{r killvar}
person$DateOfOccurrence <- NULL
observation$DateOfOccurrence <- NULL
rm(individual)
```

Now we take a look at our new data frames, one at a time.

### The *person* data frame

Looks like this:

```{r lookatperson}
summary(person)
```

We can  easily see that the variables *Start_blah* and *End_blah* are constant (= 0) and can be removed. Further, *Id*, *Source*, and *Value_Id_C* are not interesting for our purposes

```{r removefromperson}
person <- select(person, Id_I, Type, Value, Day, Month, Year)
head(person)
```

Good so far, but let us convert the three variables *Day*, *Month*, and *Year* into a **date**:

```{r}
person$date <- with(person, paste(Year, Month, Day, sep = "-")) %>%
               as.Date("%Y-%m-%d")
person$Day <- person$Month <- person$Year <- NULL
##person$Id_I <- as.character(person$Id_I)
##person$Type <- as.factor(person$Type)
str(person)
```

Which Types has dates?

```{r tabletypevalue}
with(person, table(Type, is.na(date)))
```

We note that for *BIRTH_DATE*, *DEATH_DATE*, the *date* is in fact the *Value*. So,

```{r makevalue}
change <- person$Type %in% c("BIRTH_DATE", "DEATH_DATE")
person$Value[change] <- person$date[change]
rm(change) # Just a temporary variable ...
person$date <- NULL # Not needed any more
head(person)
```
 Good reduction, but *person* is still *un-tidy*: The levels of *Type*, `r levels(person$Type)` are in fact names of *variables*. We use the package *tidyr* to expand *person*. Due to bad data, we remove all individuals with duplicate *Type == "BIRTH_LOCATION"*.
 
```{r badloc}
#x <- with(person, tapply(Type == "BIRTH_LOCATION", Id_I, sum))
person <- filter(person, !(Type == "BIRTH_LOCATION" & Value == "Locality"))
x <- group_by(person, Id_I) %>% summarize(count = sum(Type == "BIRTH_LOCATION"))
select(x, count) %>% table()
```

I do not want to complicate things here, so I keep only those with exactly one *BIRTH_LOCATION*:
```{r keeponly}
id.in <- x$Id_I[x$count == 1]
person <- filter(person, Id_I %in% id.in)
rm(id.in, x) # Temporary variables of no use anymore
```

And now we call *spread*.
 
```{r usetidyr, eval = TRUE}
library(tidyr)

person <- spread(person, Type, Value)
person$BIRTH_DATE <- as.Date(as.numeric(person$BIRTH_DATE), origin = "1970-01-01")
person$DEATH_DATE <- as.Date(as.numeric(person$DEATH_DATE), origin = "1970-01-01")
head(person)
```

Dates are now fixed. Due to the earlier un-tidiness, variables are all *character*, which we must fix.

```{r trimma}
library(stringr)
person$BIRTH_LOCATION <- str_trim(person$BIRTH_LOCATION)
person <- filter(person, BIRTH_LOCATION %in%  
                  c("SKELLEFTEÅ", "JÖRN", "NORSJÖ", "BURTRÄSK", "LÖVÅNGER"))
person$MULTIPLE_BIRTH <- str_trim(person$MULTIPLE_BIRTH)
person$MULTIPLE_BIRTH <- factor(person$MULTIPLE_BIRTH, labels = c("2", "3", "1"))
person$MULTIPLE_BIRTH <- relevel(person$MULTIPLE_BIRTH, ref = "1")
person <- filter(person, !is.na(BIRTH_DATE))
person$BIRTH_LOCATION <- factor(person$BIRTH_LOCATION)
person$SEX <- factor(person$SEX)
person$LEGITIMACY <- factor(person$LEGITIMACY)
person$Id_I <- as.character(person$Id_I)
summary(person)
```

## The *observation* data frame

We first remove all individuals not present in the *person* data frame.

```{r remove}
observation <- filter(observation, Id_I %in% person$Id_I)
summary(observation)
```

We start by cleaning this a bit, make dates, remove not-so-interesting variables, etc.

```{r cleanobs}
observation$date <- with(observation, paste(Year, Month, Day, sep = "-")) %>%
    as.Date("%Y-%m-%d")
observation$start_date <- with(observation, 
                               paste(Start_year, Start_month, Start_year, sep = "-")) %>%
                          as.Date("%Y-%m-%d")
observation$end_date <- with(observation, paste(End_year, End_month, End_day, sep = "-")) %>%
    as.Date("%Y-%m-%d")
observation <- select(observation, Id_I, Type, Value, date, start_date, end_date)
head(observation)
id1 <- length(unique(observation$Id_I))
```

The number of individuals is `r id1`.
There is a lot of missing dates, and we need at least *one* date for each record, so let us remove all records with no dates.

```{r keepdates}
observation <- filter(observation, !(is.na(date) & is.na(start_date) & is.na(end_date)))
observation <- arrange(observation, Id_I, date, start_date)
knitr::kable(head(observation))
id2 <- length(unique(observation$Id_I))
```

Now the number of persons left is `r id2`, so we lost `r id1 - id2` of them. We also note in passing that *OCCUPATION_HISCO* is duplicated for *Id_I = 10000250* (exactly the same dates).
Makes me feel that I maybe leave occupation (HISCO) out for the course, and only concentrate on *MARRIAGE_DATE* as a time-varying covariate.

## The *INDIV_INDIV* data

```{r sumind}
indivIndiv <- read.table("rawdata/INDIV_INDIV.txt")
summary(indivIndiv)
```

OK, variables *V7:V17* are all zero, and V2 and V5 are both  constant, and V1 is just a row number, so we remove them.

```{r v1v6}
indivIndiv <- select(indivIndiv, V3, V4, V6)
names(indivIndiv) <- c("id_1", "id_2", "Relation")
head(indivIndiv)
```

Now, we are only interested of the *Mother-Child* relation, so we choose only rows with *Relation = Mother*.

```{r relmoth}
indivIndiv <- filter(indivIndiv, Relation == "Mother") %>%
              select(id_1, id_2)
names(indivIndiv) <- c("Mother", "Child")
indivIndiv <- arrange(indivIndiv, Mother, Child)
indivIndiv$Mother <- as.character(indivIndiv$Mother)
indivIndiv$Child <- as.character(indivIndiv$Child)
head(indivIndiv)
```

Now, this is a *time-fixed* relation, so we can put on the id of *Mother* directly on *person*.

```{r mothonchild}
person <- left_join(person, indivIndiv, by = c("Id_I" = "Child"), copy = TRUE)
head(person)
```



## References
